---
title: "Homework Assignment 1"
author: "Sanchi Goel"
date: 'Assigned: Oct 24, 2020, Due Sun Nov 01, 2020 11:59PM'
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
---
### Data frame basics

```{r}
nyc <- read.csv('nycflights.csv')
nyc$carrier <- as.factor(nyc$carrier)
nyc$tailnum <- as.factor(nyc$tailnum)
nyc$origin <- as.factor(nyc$origin)
nyc$dest <- as.factor(nyc$dest)
nyc$X <- NULL
str(nyc)
```

### Data Exploration

There are `r length(levels(nyc$carrier))` airlines

Following is a list: `r levels(nyc$carrier)`

No. of airlines with the code'OO' are: `r nrow(nyc[nyc$carrier=="OO",])`

The shortest flight out of any NYC airport is `r min(nyc$air_time, na.rm=TRUE)`

No. of flights with code "UA" on Jan 12th 2013 were `r nrow(nyc[nyc$carrier=="UA" & nyc$year=="2013" & nyc$month=="1" & nyc$day=="12", ])`

### Arrival Delay

Average arrival delay for all airports and all airlines combined in Jan 2013 was `r delay <- nyc[nyc$month=="1" & nyc$year == "2013",]
round(mean(delay$arr_delay, na.rm=TRUE),2) 
` minutes.

The median arrival delay for all airports and all airlines combined in Jan 2013 was `r median(delay$arr_delay, na.rm=TRUE)` minutes.

Most flights arrive earlier than expected however some flights have a large delay leading to the positive mean. 


### Airline Performance


#### Average arrival delays by airlines: 
```{r}
delayair <- round(tapply(nyc$arr_delay, nyc$carrier, mean, na.rm=TRUE),2)
delayair
```


#### BarPlot of Arrival Delays
```{r}
barplot((tapply(nyc$arr_delay, nyc$carrier, mean, na.rm=TRUE))
)
```

`r names(delayair[delayair==max(delayair)])` has the highest average delay of `r max(tapply(nyc$arr_delay, nyc$carrier, mean, na.rm=TRUE))` minutes. 

`r names(delayair[delayair==min(delayair)])` has the smallest average arrival delay of `r min(tapply(nyc$arr_delay, nyc$carrier, mean, na.rm=TRUE))` minutes. 

Both AS and HA have negative arrival delays meaning they arrive early on average. 


### Air Gain
```{r}
nyc$airgain <- nyc$dep_delay - nyc$arr_delay

```

#### Descriptive Statistics of Airgain
```{r}
summary(nyc$airgain)
```

#### Summary Graphics
```{r}
hist(nyc$airgain)
```

Airlines gain 5.6 minutes on average. 


#### Average Airgain for Airlines:
```{r}
tapply(nyc$airgain, nyc$carrier, mean, na.rm=TRUE)
```

F9 and FL do the worst job whereas HA and AS do the best. 


### Merging Data Frames

#### Data Cleaning 
```{r}
planes <- read.csv('planes.csv')
planes$X <- NULL
planes$tailnum <- as.factor(planes$tailnum)
planes$type <- as.factor(planes$type)
planes$engine <- as.factor(planes$engine)
planes$model <- as.factor(planes$model)
planes$year <- as.factor(planes$year)
planes$seats <- as.factor(planes$seats)
planes$engines <- as.factor(planes$engines)
planes$manufacturer <- as.factor(planes$manufacturer)
str(planes)

```

#### Merging Data Frames
```{r}
merged <- merge(nyc, planes, by.x="tailnum", by.y = "tailnum", sort=TRUE)
str(merged)

```

### Interesting Questions

1. Who are the top 10 manufacturers of planes that fly out of NYC airports?
```{r}
library(plyr)
man_freq <- count(merged, 'manufacturer')
man_freq <- man_freq[order(-man_freq$freq),]
man_freq[1:10,]

```


2. Is a particular carrier associated more with the most popular manufacturer Boeing? 
```{r}
boeing <- merged[merged$manufacturer=="BOEING",]
boeing_carrier <- count(boeing, 'carrier')
boeing_carrier <- boeing_carrier[order(-boeing_carrier$freq),]
boeing_carrier

```
United Airlines employs over half of the Boeing aircrafts flying out of NYC. 


3. Does the age of the aircraft have any relationship with average distance travelled? 
```{r}
barplot(tapply(merged$distance, merged$year.y, mean, na.rm=TRUE))

```

We can clearly see that the maximum average distances are covered by older planes (1990 and older)


4. Does the number of engines affect the distance covered? 
```{r}
tapply(merged$distance, merged$engines, median, na.rm=TRUE)

```
Median distance covered sharply decreases as the number of engines increase. 


5. What is the most common engine-type for planes flying out of NYC?
```{r}
table(merged$engine)
```
The most common engine is Turbo-fan. 
